---
title: "HOX depencies for AML cell lines in CCLE/depmap"
author: Fiorella Schischlik
date: October, 29th, 2021
output:
  md_document:
    variant: gfm
---

Load libraries.
```{r}
library(ggplot2)
library(ggrepel)
library(ComplexHeatmap)
library(ggpubr)
library(poolr)
library(reshape2)
library(data.table)
```

Set directories.
```{r}
# dir_depmap <- "~/OneDrive - National Institutes of Health/DepMap/"
dir_depmap <- "~/datasets/depmap/"
dir_out <- "~/src/aml-2019/results/HOX_dependency/"
dir_resouces <-"~/src/aml-2019/resources/"
# on biowulf
#dir_depmap <- "/data/Lab_ruppin/projects/sl-nf1-cdkn2a/processed_data/"
#dir_out <- "/data/Lab_ruppin/schischlikf2/aml-2019/"
```

Load data.
```{r}

# load depmap data
depmap <- readRDS(paste(dir_depmap, "depmap_21Q3_screen_data.rds", sep=""))

```

Define HOX genes.
```{r}
# Resources: https://www.haematologica.org/article/view/9926
# As for the HOX question, we can divide leukemia (not just AML, but leukemia 
# – AML and ALL, so we have more) cell lines into HOX positive and HOX negative 
# to do the same figure as above. Here we can do a statistic between HOX 
# positive compared to HOX negative. To define HOX positive and negative,  The 
# list of HOX genes activated in AML cell lines are mostly of the HOXA cluster 
# (HOXA5, HOXA7, HOXA9, HOXA10, HOXA11) or the HOXB cluster (HOXB3, HOXB4, 
# HOXB5, HOXB6, HOXB7). So the HOX positive cell lines would be either HOXA 
# OR HOXB positive (both clubbed into the HOX positive group). Any cell line 
# that has neither HOXA nor HOXB expressed would be the HOX negative group.

HOXA <- c("HOXA1", "HOXA2", "HOXA3", "HOXA4", 
          "HOXA5", "HOXA6", "HOXA7", "HOXA9", 
          "HOXA10", "HOXA11", "HOXA13")
HOXB <- c("HOXB1", "HOXB2", "HOXB3", "HOXB4", 
          "HOXB5", "HOXB6", "HOXB7", "HOXB8",
          "HOXB9", "HOXB13")

# Other HOX related genes
add_genes <- c("MEIS1", "MEIS2", "PBX1")

HOX <- c(HOXA, HOXB, add_genes)

gene_signatures <- list(
  HOXA = HOXA,
  HOXB = HOXB,
  HOX = HOX)

```

Add AML manual HOX annotation.
```{r}

# Ani Deshpande annotation
aml_manual_annotation <- readxl::read_excel(paste0(dir_resouces, "DepMap_AML_Cell_Lines_Hox_KMT2A_Status_etc.xlsx"))

# Annotate which cell lines have expression and crispr data
dp_cls <- colnames(depmap$mrna)
crispr_cls <- dp_cls[colSums(depmap$gene_fitness_crispr_depmap_21q3, na.rm=TRUE) != 0]
rnai_cls <- dp_cls[colSums(depmap$gene_fitness_rnai_depmap, na.rm=TRUE) != 0]

# which one's do overlap with the aml cell lines ?
aml_manual_annotation$have_crispr <- aml_manual_annotation$DepMap_ID %in% crispr_cls
aml_manual_annotation$have_rnai <- aml_manual_annotation$DepMap_ID %in% rnai_cls

```

Define gene module overall expression score.
```{r}

subset_object <- function(
  data_object, 
  meta_selection, 
  meta_id="DepMap_ID"){
  
  new_data_object <- list()
  new_data_object$genes <- data_object$genes
  new_data_object$meta_data <- data_object$meta_data[data_object$meta_data[[meta_id]] %in% meta_selection, ]
  new_data_object$mrna <- data_object$mrna[, meta_selection]
  new_data_object$scna <- data_object$snca[, meta_selection]
  new_data_object$gene_fitness_crispr_depmap_21q3 <- data_object$gene_fitness_crispr_depmap_21q3[, meta_selection]
  new_data_object$gene_fitness_rnai_depmap <- data_object$gene_fitness_rnai_depmap[, meta_selection]
  
  return(new_data_object)
  
}

# Modified from Livnat et al.

discretize <- function(v, n_cat){
  
  q1 <- quantile(v, seq(from = (1/n_cat), to = 1, by = (1/n_cat)), na.rm = TRUE)
  u <- matrix(nrow = length(v))
  
  for(i in 2:n_cat){
    u[(v >= q1[i-1]) & (v < q1[i])] <- i
  }
  
  return(u)
}

get_semi_random_OE <- function(
  object,
  genes_dist_q,
  b_signature,
  num_rounds = 1000,
  full_flag = FALSE){
 
  sign_q <- as.matrix(table(genes_dist_q[b_signature]))
  q <- rownames(sign_q)
  idx_all <- c()
  B <- matrix(data = FALSE, nrow = length(genes_dist_q), ncol = num_rounds)
  Q <- matrix(data = 0, nrow = length(genes_dist_q), ncol = num_rounds)
  
  for (i in 1:nrow(sign_q)){
    num_genes <- sign_q[i]
    if(num_genes > 0){
      idx <- which(is.element(genes_dist_q, q[i]))
      for (j in 1:num_rounds){
        idxj <- sample(idx, num_genes) 
        Q[i,j] <- sum(B[idxj,j]==TRUE)
        B[idxj,j] <- TRUE
      }  
    }
  }
  rand_scores <- apply(B, 2, function(x) colMeans(object$zscores[x,]))
  if(full_flag){return(rand_scores)}
  rand_scores <- rowMeans(rand_scores)
  return(rand_scores)
}

get_overall_expression <- function(
  object,
  gene_signature = NULL,  # list object
  num_rounds = 1000,
  full_flag = FALSE){
  
  set.seed(42)
  object$genes_mean <- rowMeans(object$mrna, na.rm = TRUE)
  object$zscores <- sweep(object$mrna, 1, object$genes_mean, FUN = '-')
  object$genes_dist <- object$genes_mean
  
  # Bin genes by expression levels
  object$genes_dist_q <- discretize(object$genes_dist , n_cat = 50)
  
  # Create sig_scores matrix
  object$sig_scores <- matrix(data = 0, nrow = ncol(object$mrna),
                              ncol = length(gene_signature))
  colnames(object$sig_scores) <- names(gene_signature)
  object$sig_scores_raw <- object$sig_scores
  
  rand_flag <- is.null(object$rand_scores) | !all(is.element(names(gene_signature), colnames(object$rand_scores)))
  
  if(rand_flag){
    print("Computing also random scores.")
    object$rand_scores <- object$sig_scores
  }
  
  for (signature in names(gene_signature)){
    
    b_signature <- is.element(object$genes, gene_signature[[signature]])
    
    if(sum(b_signature) < 2){ next() }
    
    if(rand_flag){
      rand_scores <- get_semi_random_OE(
        object, 
        object$genes_dist_q, 
        b_signature, 
        num_rounds = num_rounds)
    }else{
      rand_scores <- object$rand_scores[, signature] }
    
    raw_scores <- colMeans(object$zscores[b_signature,])
    final_scores <- raw_scores - rand_scores
    object$sig_scores[ ,signature] <- final_scores
    object$sig_scores_raw[ ,signature] <- raw_scores
    object$rand_scores[ ,signature] <- rand_scores
  }
  
  if(full_flag){ return(object) }
  
  sig_scores <- object$sig_scores
  
  return(sig_scores)
}

```

Select and annotate AML cell lines with HOX module score.
```{r}

aml_cell_lines <- rownames(depmap$meta_data[grep("AML", depmap$meta_data$subtype),])
aml_depmap <- subset_object(depmap, meta_selection = aml_cell_lines)
hox_scores <- get_overall_expression(aml_depmap, gene_signature = gene_signatures)

```

Select and annotate ALL cell lines with HOX module score.
```{r}

all_cell_lines <- rownames(depmap$meta_data[grep("ALL", depmap$meta_data$subtype),])
all_depmap <- subset_object(depmap, meta_selection = all_cell_lines)
all_hox_scores <- get_overall_expression(all_depmap, gene_signature = gene_signatures)

```

Visualize HOX genes (for AML).
```{r}

# check HOX gene distribution in aml_cell_lines, witout NA's
aml_cell_lines <- rownames(depmap$meta_data[grep("AML", depmap$meta_data$subtype),])
hox_expr <- depmap$mrna[HOX, aml_cell_lines]
colnames(hox_expr) <- depmap$meta_data[aml_cell_lines, ]$cell_line
hox_expr_naomit <- t(na.omit(t(hox_expr)))

# subselect aml manual annotation
aml_manual_annotation_sub <- aml_manual_annotation[aml_manual_annotation$stripped_cell_line_name %in% colnames(hox_expr_naomit), ]

aml_manual_annotation_sub[is.na(aml_manual_annotation_sub$Hox_High_Known), ]$Hox_High_Known <- "No"
aml_manual_annotation_sub$stripped_cell_line_name <- factor(
  aml_manual_annotation_sub$stripped_cell_line_name, levels=colnames(hox_expr_naomit))
aml_manual_annotation_sub <- aml_manual_annotation_sub[order(aml_manual_annotation_sub$stripped_cell_line_name), ]

# Add annotation
ha <- HeatmapAnnotation(
  HOX_positive = aml_manual_annotation_sub$Hox_High_Known,
  have_crispr = aml_manual_annotation_sub$have_crispr,
  have_rnai = aml_manual_annotation_sub$have_rnai,
  gp = gpar(col = "white"))

# Plot 
Heatmap(
  t(scale(t(hox_expr_naomit))), 
  top_annotation = ha,
  heatmap_legend_param = list(title = "Expression \n z-score"),
  cluster_rows = FALSE)

```

Visualize HOX genes (for ALL).
```{r}

# check HOX gene distribution in aml_cell_lines, without NA's
all_cell_lines <- rownames(depmap$meta_data[grep("ALL", depmap$meta_data$subtype),])
hox_expr_all <- depmap$mrna[HOX, all_cell_lines]
colnames(hox_expr_all) <- depmap$meta_data[all_cell_lines, ]$cell_line
hox_expr_all_naomit <- t(na.omit(t(hox_expr_all)))

# Add annotation
ha <- HeatmapAnnotation(
  HOX = na.omit(all_hox_scores[,"HOX"]),
  HOXA = na.omit(all_hox_scores[,"HOXA"]),
  HOXB = na.omit(all_hox_scores[,"HOXB"]),
  gp = gpar(col = "white"))

# Plot 
Heatmap(
  t(scale(t(hox_expr_all_naomit))), 
  top_annotation = ha,
  heatmap_legend_param = list(title = "Expression \n z-score"),
  cluster_rows=FALSE)

```

Plot AML and ALL together.
```{r}

# Scale expression
hox <- cbind(hox_expr_all_naomit, hox_expr_naomit)
t(scale(t(hox)))
pca_on_hox <- prcomp(hox)
# 4 clusters
fit <- kmeans(pca_on_hox$rotation, 4)

# Annotation
ha <- HeatmapAnnotation(
  cell_line = c(rep("ALL", dim(hox_expr_all_naomit)[2]), rep("AML", dim(hox_expr_naomit)[2])),
  gp = gpar(col = "white"))

Heatmap(
  t(scale(t(hox))), 
  heatmap_legend_param = list(title = "Expression \n z-score"),
  top_annotation = ha, 
  cluster_rows = FALSE)

```

Identify HOX low and HOX high cell lines in AML.
```{r}

# Try adding up scores
scaled_matrix <- t(scale(t(hox_expr_naomit)))
median_score <- median(colSums((scaled_matrix > 0) * 1))
lower_t <- quantile(colSums((scaled_matrix > 0) * 1), c(0:3/3))[2]

# Try adding up sum of expression
hoxa_expr_score <- colSums(hox_expr_naomit[HOXA, ])
hoxa_score <- hoxa_expr_score < quantile(hoxa_expr_score, probs = 0.25)
hoxb_expr_score <- colSums(hox_expr_naomit[HOXB, ])
hoxb_score <- hoxb_expr_score < quantile(hoxb_expr_score, probs = 0.25)

# Add ab zscores
# I'm trying this approach so that each gene will have equal contribution
sum_zscore <- colSums(hox_expr_naomit[HOX, ])

#ha <- HeatmapAnnotation(
#  HOX_positive = aml_manual_annotation_sub$Hox_High_Known,
#  hox_all = sum_zscore,
#  hox_livnat = na.omit(hox_scores[,"HOX"]),
#  hoxa_expr_score = hoxa_expr_score,
#  hoxa_score = hoxa_score,
#  hoxb_expr_score = hoxb_expr_score,
#  hoxb_score = hoxb_score,
#  hox_zscore = as.factor((colSums((scaled_matrix > 0) * 1) > lower_t) * 1),
#  have_crispr = aml_manual_annotation_sub$have_crispr,
#  have_rnai = aml_manual_annotation_sub$have_rnai,
#  gp = gpar(col = "white"))

ha <- HeatmapAnnotation(
  HOX_positive = aml_manual_annotation_sub$Hox_High_Known,
  HOX = na.omit(hox_scores[,"HOX"]),
  HOXA = na.omit(hox_scores[,"HOXA"]),
  HOXB = na.omit(hox_scores[,"HOXB"]),
  have_crispr = aml_manual_annotation_sub$have_crispr,
  have_rnai = aml_manual_annotation_sub$have_rnai,
  gp = gpar(col = "white"))

Heatmap(
  t(scale(t(hox_expr_naomit))), 
  heatmap_legend_param = list(title = "Expression \n z-score"),
  top_annotation = ha, 
  cluster_rows = FALSE)

```

Linear regression function.
```{r}

run_lm <- function(
  aml_depmap,
  model = "gene_fitness ~ hox_oe", 
  variable_of_interest = "hox_oe",
  goi = aml_depmap$genes,
  slot='gene_fitness_crispr_depmap_21q3'){
  
  collect_results <- data.frame(
  Estimate = numeric(),
  Std.Error = numeric(),
  t.value = numeric(), 
  p.value = numeric(),
  gene = character())

  for (gene in goi){
  
    #print(gene)
    data_lm <- data.frame(
      cell_line=aml_depmap$meta_data$cell_line,
      hox_oe=aml_depmap$meta_data$hox_oe,
      hoxa_oe=aml_depmap$meta_data$hoxa_oe,
      hoxb_oe=aml_depmap$meta_data$hoxb_oe,
      gene_fitness=aml_depmap[[slot]][gene, ])
  
    if (sum(!is.na(data_lm$gene_fitness)) > 10){
      # Run linear model
      results <- lm(model, data = data_lm)
  
      collect_results <- rbind(
        collect_results,
        c(summary(results)$coefficients[variable_of_interest, ], gene=gene))
    }else{
      collect_results <- rbind(collect_results, c(NA, NA, NA, NA, gene))
    }
  }

  colnames(collect_results) <- c("Estimate", "Std.Error", 
                            "t.value", "p.value", "gene")
  
  return(collect_results)
}

```

Run linear regression model for AML.
```{r}

goi <- c("SGF29", "ENY2", "MLLT1", "DOT1L", "KMT2A", "CSNK2A1", "CSNK2A2",
         "CSNK2B", "HOXA9", "MEIS1", "KAT7", "MLLT10", "JADE3", "AFF2", 
         "BRD4", "MLLT3", "AFF4", "CDK9", "CDK7", "MEN1")

# Add HOX expression etc to meta dataframe
aml_depmap$meta_data$hox_oe <- hox_scores[,"HOX"]
aml_depmap$meta_data$hoxa_oe <- hox_scores[,"HOXA"]
aml_depmap$meta_data$hoxb_oe <- hox_scores[,"HOXB"]

# crispr
hox_oe_results <- run_lm(
  aml_depmap, 
  model="gene_fitness ~ hox_oe",
  variable_of_interest = "hox_oe", 
  slot='gene_fitness_crispr_depmap_21q3')

hoxa_oe_results <- run_lm(
  aml_depmap, model="gene_fitness ~ hoxa_oe",
  variable_of_interest = "hoxa_oe")

hoxb_oe_results <- run_lm(
  aml_depmap, model="gene_fitness ~ hoxb_oe",
  variable_of_interest = "hoxb_oe")

# rnai
hox_oe_results_rnai <- run_lm(
  aml_depmap, model="gene_fitness ~ hox_oe",
  variable_of_interest = "hox_oe",
  slot='gene_fitness_rnai_depmap')

hoxa_oe_results_rnai <- run_lm(
  aml_depmap, model="gene_fitness ~ hoxa_oe",
  variable_of_interest = "hoxa_oe",
  slot='gene_fitness_rnai_depmap')

hoxb_oe_results_rnai <- run_lm(
  aml_depmap, model="gene_fitness ~ hoxb_oe",
  variable_of_interest = "hoxb_oe",
  slot='gene_fitness_rnai_depmap')

```

Run linear regression model for ALL.
```{r}

# Add HOX expression etc to meta dataframe
all_depmap$meta_data$hox_oe <- all_hox_scores[,"HOX"]
all_depmap$meta_data$hoxa_oe <- all_hox_scores[,"HOXA"]
all_depmap$meta_data$hoxb_oe <- all_hox_scores[,"HOXB"]

# crispr
hox_oe_results <- run_lm(
  all_depmap, 
  model="gene_fitness ~ hox_oe",
  variable_of_interest = "hox_oe", 
  slot='gene_fitness_crispr_depmap_21q3')

hoxa_oe_results <- run_lm(
  all_depmap, 
  model="gene_fitness ~ hoxa_oe",
  variable_of_interest = "hoxa_oe", 
  slot='gene_fitness_crispr_depmap_21q3')

hoxb_oe_results <- run_lm(
  all_depmap, 
  model="gene_fitness ~ hoxb_oe",
  variable_of_interest = "hoxb_oe", 
  slot='gene_fitness_crispr_depmap_21q3')

```

Annotate results (AML).
```{r}

common_essentials <- read.csv("/Users/schischlikf/Google Drive/My Drive/NIH_NCI/src/aml-2019/resources/CRISPR_common_essentials.csv")

genes <- as.character(sapply(common_essentials$gene, function(x) strsplit(x, split=" ")[[1]][1]))

hoxa_oe_results$common_essentials <- FALSE
hoxa_oe_results[hoxa_oe_results$gene %in% genes, ]$common_essentials <- TRUE

hoxa_oe_results$drugbank <- sapply(hoxa_oe_results$gene, FUN=function(x) annotate.drugs.drugbank(x, drug.list=drugs))
hoxa_oe_results$FDR <- p.adjust(hoxa_oe_results$p.value, method="fdr")

write.csv(hoxa_oe_results, file=paste(dir_out, "HOXA_dependency_linear_regression_results.tsv"))

```

Visualize with vulcano plot.
```{r}

library(ggrepel)

hoxb_oe_results$Estimate <- as.numeric(as.character(hoxb_oe_results$Estimate))
hoxb_oe_results$p.value <- as.numeric(as.character(hoxb_oe_results$p.value))

hoxb_oe_results$HOX_reg <- FALSE
hoxb_oe_results[hoxb_oe_results$gene %in% goi, ]$HOX_reg <- TRUE

cutoffpvaladj <- 0.05

# Plot vulcano
g <- ggplot(data=hoxb_oe_results, aes(x=Estimate, y=-log10(p.value)))
g <- g + geom_vline(xintercept = 0, linetype="dashed", color="grey", size=0.7)
g <- g + geom_hline(yintercept = -log10(cutoffpvaladj), linetype="dashed", color="grey", size=0.7)
g <- g + geom_point(aes(x=Estimate, y=-log10(p.value), fill=HOX_reg, 
                        color=HOX_reg), pch=21, size=2, alpha=0.5)
g <- g + geom_point(data=dplyr::filter(hoxb_oe_results, HOX_reg==TRUE),
                    aes(x=Estimate, y=-log10(p.value)), 
                    fill="#e41a1c", pch=21, size=2, alpha=0.9)
g <- g + scale_fill_manual(values=c("blue", "#e41a1c"))
g <- g + scale_color_manual(values=c("blue","#e41a1c"))
g <- g + geom_text_repel(data=dplyr::filter(hoxb_oe_results, HOX_reg==TRUE), 
  aes(x=Estimate, y=-log10(p.value), label=gene),
  arrow = arrow(length = unit(0.02, "npc"), type = "closed", ends = "first"),
  segment.color = "grey50", size=3)
g <- g + xlab("Estimate")
g <- g + ylab("-Log10(P value)")
g <- g + theme_bw()
g <- g + theme(
  aspect.ratio=1,
  axis.text = element_text(size = 12))
plot(g)

pdf(width=5, height=5, 
    file=paste(args$dir_results, "figure_hox_vulcano", mode, ".pdf", sep=""), 
    useDingbats = FALSE)
plot(g)
dev.off()

```

Identify HOX low and HOX high cell lines in ALL
```{r}

scaled_matrix <- t(scale(t(hox_expr_all_naomit)))
median_score <- median(colSums((scaled_matrix > 0) * 1))
lower_t <- quantile(colSums((scaled_matrix > 0) * 1), c(0:3/3))[2]

ha <- HeatmapAnnotation(
  hox = as.factor((colSums((scaled_matrix > 0) * 1) > lower_t) * 1),
  gp = gpar(col = "white"))

Heatmap(
  t(scale(t(hox_expr_all_naomit))), 
  heatmap_legend_param = list(title = "Expression \n z-score"),
  top_annotation = ha, 
  cluster_rows = FALSE)

```

Essentiality function.
```{r}

essentiality <- function(
  screen=screen_data,
  sl_pair=sl_pair,
  data_slots=c("mrna", "scna"),
  gene_fitness_slots=c("crispr_depmap_21q3", "rnai_depmap"),
  q2=FALSE){
  
  # results columns
  results <- c()
  #results_column <- c("SL_A_B_", "SDL_A_B_", "SL_B_A_", "SR_B_A_")
  #results_column <- as.character(sapply(data_slots, function(x) paste(results_column, x, sep="")))
  #results_column <- append(c('GENE_A', 'GENE_B'), results_column)
  #names(results) <- results_column
  
  # Just in case make sure the sl_pair does not contain NA values
  if(sum(is.na(sl_pair)) == 0){
    
      if (q2){
        
        # q2 slots
        q2_slots <- paste0(data_slots, "_q2")
        
      for (score in gene_fitness_slots){
        
        for (slot in q2_slots){
          
          results[paste("GENE_A_", score, "_fit_score#cl")] <- sum((!is.na(screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_A"], ])) * 1)
        results[paste("GENE_B_", score, "_fit_score#cl")] <- sum((!is.na(screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_B"], ])) * 1)
  
          molecdata_q2 <- screen[[slot]][sl_pair["GENE_A"],]
          
          #print(molecdata_q2)
          
          results[paste("SL_A_B", slot, score, sep="_")] <- wilcox.test.na(
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_B"], molecdata_q2 == 0],
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_B"], molecdata_q2 > 0],
            alternative1 = "less")
          
          results[paste("SDL_A_B", slot, score, sep="_")] <- wilcox.test.na(
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_B"], molecdata_q2 < 2],
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_B"], molecdata_q2 == 2],
            alternative1 = "greater")
          
          # Change direction B -> A
          molecdata_q2 <- screen[[slot]][sl_pair["GENE_B"],]
          
          results[paste("SL_B_A", slot, score, sep="_")] <- wilcox.test.na(
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_A"], molecdata_q2 == 0],
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_A"], molecdata_q2 > 0],
            alternative1 = "less")
          
          results[paste("SR_B_A_", slot, score, sep="_")] <- wilcox.test.na(
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_A"], molecdata_q2 < 2],
            screen[[paste0("gene_fitness_",score)]][sl_pair["GENE_A"], molecdata_q2 == 2],
            alternative1 = "greater")
          }
        }
      }
  }
    return(results)
}



wilcox.test.na = function(x,y, alternative1, paired=FALSE) {
  
  tryCatch(
    wilcox.test(x, y, alternative=alternative1, paired=paired)$p.value,
    error = function(e) NA)
  
}

```

Run essentiality. 
```{r}

run_essentiality <- function(
  screen=depmap,
  data){
  
  # Define HOX_status
  HOX_status <- (hoxa_expr_score < quantile(hoxa_expr_score, probs = 0.25)) * 1
  # HOX_status <- (colSums((scaled_matrix > 0) * 1) > lower_t) * 1
  
  # Get expression
  expression <- depmap$mrna
  colnames(expression) <- depmap$meta_data$cell_line
  cell_lines_symbol <- colnames(data)
  cls_matrix <- expression[, cell_lines_symbol]
  cls_matrix <- rbind(HOX_status=c(HOX_status * 2), cls_matrix)

  # crispr
  essentiality_dep <- depmap$gene_fitness_crispr_depmap_21q3
  colnames(essentiality_dep) <- depmap$meta_data$cell_line
  matrix_ess <- essentiality_dep[, cell_lines_symbol]
  matrix_ess <- rbind(HOX_status=c(HOX_status * 2), matrix_ess)

  # rnai
  rnai <- depmap$gene_fitness_rnai_depmap
  colnames(rnai) <- depmap$meta_data$cell_line
  matrix_rnai <- rnai[, cell_lines_symbol]
  matrix_rnai <- rbind(HOX_status=c(HOX_status * 2), matrix_rnai)

  screen <- list(
    mrna_q2=cls_matrix,
    gene_fitness_crispr_depmap_21q3=matrix_ess,
    gene_fitness_rnai_depmap=matrix_rnai
  )

  gene_a <- "HOX_status"
  gene_names <- row.names(screen$mrna)
  
  essentiality_multiple <- as.data.frame(t(
    as.data.frame(
      sapply(gene_names, function(x) essentiality(
        screen=screen,
        sl_pair=c(GENE_A = gene_a, GENE_B = x),
        data_slots=c("mrna"),
        gene_fitness_slots=c("crispr_depmap_21q3", "rnai_depmap"),
        q2=TRUE)))))
  
  return(essentiality_multiple)
}

aml_essentiality <- run_essentiality(
  screen=depmap,
  data=hox_expr_naomit)

all_essentiality <- run_essentiality(
  screen=depmap,
  data=hox_expr_all_naomit)

```

Add FDR.
```{r}

essentiality_multiple <- data.frame(aml_essentiality)
essentiality_multiple$SDL_BH_crispr <- p.adjust(essentiality_multiple$SDL_A_B_mrna_q2_crispr_depmap_21q3, method="fdr")
essentiality_multiple$SDL_BH_rnai <- p.adjust(essentiality_multiple$SDL_A_B_mrna_q2_rnai_depmap, method="fdr")
essentiality_multiple <- essentiality_multiple[order(essentiality_multiple$SDL_A_B_mrna_q2_crispr_depmap_21q3), ]

sdl <- data.frame(na.omit(essentiality_multiple[c(
  "SDL_A_B_mrna_q2_crispr_depmap_21q3", "SDL_BH_crispr",
  "SDL_A_B_mrna_q2_rnai_depmap", "SDL_BH_rnai")]))

# sdl$fisher <- apply(
#   sdl, 
#   1, 
#   function(x) { (poolr::fisher(x, adjust = "none"))$p})
# 
# sdl$BH <- p.adjust(sdl$fisher, method="BH")

```

Add drug targets.
```{r}

sdl$gene_name <- row.names(sdl)
drugs <- read.csv("~/Google Drive/My Drive/NIH_NCI/datasets/Druggability/Druggability.tsv", header=TRUE, sep="\t")

annotate.drugs.drugbank <- function(gene_name, drug.list=drugs){
  #print(gene_name)
  #print(gene_name)
  #print(head(drug.list))
  drug_gene <- drug.list[drug.list$gene_name==as.character(gene_name),]
  if(!(is.data.frame(drug_gene) && nrow(drug_gene)==0)){
    matched.drugs <- paste(unique(drug_gene$drug_name), collapse=",")
  }
  else{
    matched.drugs <- ""
  }
  #print(matched.drugs)
  return(matched.drugs)
}

sdl$drugbank <- sapply(sdl$gene_name, FUN=function(x) annotate.drugs.drugbank(x, drug.list=drugs))

write.csv(sdl, file=paste(dir_out, "HOX_dependency_ALL.tsv"))

# only subselect genes with drug targets
sdl_sub <-sdl[sdl$drugbank!="", ]
sdl_sub$BH_sub <- p.adjust(sdl_sub$SDL_A_B_mrna_q2_crispr_depmap_21q3, method="BH")

```

Prepare data for plotting.
```{r}

get_data_in_format <- function(
  depmap, 
  data,
  HOX_status){
  
  # Expression
  expression <- depmap$mrna
  colnames(expression) <- depmap$meta_data$cell_line
  cell_lines_symbol <- colnames(data)
  cls_matrix <- expression[, cell_lines_symbol]
  cls_matrix <- rbind(HOX_status=HOX_status, cls_matrix)
  
  # crispr
  essentiality_dep <- depmap$gene_fitness_crispr_depmap_21q3
  colnames(essentiality_dep) <- depmap$meta_data$cell_line
  matrix_ess <- essentiality_dep[, cell_lines_symbol]
  matrix_ess <- rbind(HOX_status=HOX_status, matrix_ess)

  # rnai
  rnai <- depmap$gene_fitness_rnai_depmap
  colnames(rnai) <- depmap$meta_data$cell_line
  matrix_rnai <- rnai[, cell_lines_symbol]
  matrix_rnai <- rbind(HOX_status=HOX_status, matrix_rnai)


  screen <- list(
    mrna_q2=cls_matrix,
    gene_fitness_crispr_depmap_21q3=matrix_ess,
    gene_fitness_rnai_depmap=matrix_rnai)

  gene_a <- "HOX_status"
  gene_names <- row.names(screen$mrna)
  
  return(screen)
}

```

Visualize the gene essentiality data. (Boxplots)
```{r}

# Select genes to plot
#gof <- row.names(head(essentiality_multiple, 10))
sdl <- sdl[order(sdl$SDL_A_B_mrna_q2_crispr_depmap_21q3), ]
gof <- c("KMT2A", "DOT1L", "MEN1")
#gof <- row.names(head(sdl, 10))

expression <- screen_to_plot$mrna_q2["HOX_status", ]
gene_fitness_score <- screen_to_plot$gene_fitness_crispr_depmap_21q3[gof, ]
#gene_fitness_score <- screen$gene_fitness_rnai_depmap[gof, ]

# Reshape data
gfs_melted <- reshape2::melt(t(gene_fitness_score))
colnames(gfs_melted) <- c("cell_line", "gene_name", "gene_fitness_score")
gfs_melted$strata <- rep(expression, length(unique(gfs_melted$gene_name)))
gfs_melted$strata_name <- "HOX pos"
gfs_melted[gfs_melted$strata==0, ]$strata_name <- "HOX neg"

#gene_ess_plot <- data.frame(
#  expression = expression,
#  gene_fitness_score=gene_fitness_score,
#  strata=expression)

my_comparisons <- list(c("HOX pos", "HOX neg"))

p <- ggboxplot(gfs_melted, x = "strata_name", y = "gene_fitness_score",
               color = "strata_name", palette =c("#00AFBB", "#E7B800"),
               add = "jitter", shape = "strata_name")
p <- p + ylim(c(-1.5, 0.8))
p <- p + stat_compare_means(
  comparisons = my_comparisons, 
  method = "wilcox.test",
  paired = FALSE)
p <- p + facet_wrap(~gene_name)
plot(p)

# Select single gene
expression <- screen$mrna_q2["HOX_status", ]
gene_fitness_score <- screen$gene_fitness_crispr_depmap_21q3["SYNE3", ]
#gene_fitness_score <- screen$gene_fitness_rnai_depmap["POLE2", ]

gene_ess_plot <- data.frame(
  expression = expression,
  gene_fitness_score=gene_fitness_score,
  strata=expression)

my_comparisons <- list( c("0", "2") )

p <- ggboxplot(gene_ess_plot, x = "strata", y = "gene_fitness_score",
               color = "strata", palette =c("#00AFBB", "#E7B800"),
               add = "jitter", shape = "strata")
p <- p + stat_compare_means(
  comparisons = my_comparisons, 
  method = "wilcox.test",
  paired = FALSE)
plot(p)


```

Run correlation.
```{r}


quick_cor <- function(gene_corr, x, y){
    
    gene_cor <- as.data.frame(t(screen_to_plot$gene_fitness_crispr_depmap_21q3[c(x, y),]))
    gene_cor_na_omit <- na.omit(gene_cor)
    gene_cor_na_omit$gene_name <- row.names(gene_cor_na_omit)
    
    #print(gene_cor_na_omit)
    
    if(dim(gene_cor_na_omit)[1] == 0){
      pearson_test <- as.list(c(p.value=NA, estimate=NA))
    }else{
      pearson_test <- cor.test(
      gene_cor_na_omit[[x]], 
      gene_cor_na_omit[[y]], 
      method="pearson", 
      alternative="less")
    }
    
    return(c(pearson_test$p.value, pearson_test$estimate))
}

run_correlation <- function(
  screen=depmap,
  data,
  HOX_status=hoxa_expr_scores){
  
  # Get expression
  expression <- depmap$mrna
  colnames(expression) <- depmap$meta_data$cell_line
  cell_lines_symbol <- colnames(data)
  cls_matrix <- expression[, cell_lines_symbol]
  cls_matrix <- rbind(HOX_status=HOX_status, cls_matrix)

  # crispr
  essentiality_dep <- depmap$gene_fitness_crispr_depmap_21q3
  colnames(essentiality_dep) <- depmap$meta_data$cell_line
  matrix_ess <- essentiality_dep[, cell_lines_symbol]
  matrix_ess <- rbind(HOX_status=HOX_status, matrix_ess)

  # rnai
  rnai <- depmap$gene_fitness_rnai_depmap
  colnames(rnai) <- depmap$meta_data$cell_line
  matrix_rnai <- rnai[, cell_lines_symbol]
  matrix_rnai <- rbind(HOX_status=HOX_status, matrix_rnai)

  screen <- list(
    mrna_q2=cls_matrix,
    gene_fitness_crispr_depmap_21q3=matrix_ess,
    gene_fitness_rnai_depmap=matrix_rnai
  )

  gene_a <- "HOX_status"
  gene_names <- row.names(screen$mrna)
  
  cor_multiple <- sapply(gene_names, function(x) quick_cor(
    screen,
    x=gene_a,
    y=x))
  
  return(cor_multiple)
}

aml_corr <- run_correlation(
  screen=depmap,
  data=hox_expr_naomit,
  HOX_status = hoxa_expr_score)

aml_corr <- data.frame(t(aml_corr))
colnames(aml_corr) <- c("pvalue", "cor")

aml_corr$FDR <- p.adjust(aml_corr$pvalue, method="fdr")
aml_corr <- aml_corr[order(aml_corr$pvalue),]

write.csv(aml_corr,
          file = paste(dir_out, "HOX_dependency_aml_specific_pearson.tsv"))
```

Visualize correlation.
```{r}

#HOX_status <- (hoxa_expr_score < quantile(hoxa_expr_score, probs = 0.25)) * 1

#screen_to_plot <- get_data_in_format(
#  depmap, 
#  data=hox_expr_naomit, 
#  HOX_status = hoxa_expr_score)

#genes_to_plot <- c("HOX_status", row.names(aml_corr)[32:48])
# genes_to_plot <- c("HOX_status", "DOT1L", "MEN1", "KMT2A", "MLLT1")

slot <- 'gene_fitness_rnai_depmap'
hox_res <- hoxa_oe_results
depmap_object <- all_depmap
expr_obj <- hox_expr_all_naomit

hox_res$t.value <- as.numeric(as.character(hox_res$t.value))
hox_res$p.value <- as.numeric(as.character(hox_res$p.value))
goi <- head(na.omit((hox_res[hox_res$t.value < -3 & hox_res$p.value < 0.001, ])), 12)
goi <- goi[order(goi$p.value), ]

genes_to_plot <- c("HOX_status", goi$gene)

HOX_score <- depmap_object$meta_data$hoxa_oe

screen_to_plot <- get_data_in_format(
  depmap=depmap_object,
  data=expr_obj,
  HOX_status = HOX_score)

gene_corr <- as.data.frame(t(screen_to_plot[[slot]][genes_to_plot,]))
gene_corr$cell_lines <- row.names(gene_corr)

to_plot <- data.table::melt(gene_corr, id.vars=c("HOX_status", "cell_lines"))
colnames(to_plot) <- c("HOX_score", "cell_lines", "gene_name", "dependency_score")
#cor.test(gene_corr$HOX_status, gene_corr$MRPS23, method="pearson", alternative="less")

g <- ggplot(data=to_plot, aes(x=HOX_score, y=dependency_score, label=cell_lines))
g <- g + geom_point()
#g <- g + ylim(c(-1, 0.5))
g <- g + stat_smooth(method = "lm", col = "red")
g <- g + geom_text_repel(size=2)
g <- g + facet_wrap(~gene_name, scales="free_y")
plot(g)

fn <-
  sprintf(
    paste(
      dir_out,
      "depmap_crispr_HOX_dependency_crispr_top_pearson.pdf",
      sep = ""
    )
  )
pdf(fn, height = 10, width = 11)
print(g)
dev.off()


```


