---
title: "List of prognostic genes from TCGA, TARGET, BEAT AML study."
author: Fiorella Schischlik
date: December 9, 2019
output:
  md_document:
    variant: gfm
---

Resources and other issues
```{r}
Resources
# https://www.biostars.org/p/153013/
Issues
# Should I standardize age before cox ?
```

Load libraries.
```{r}

library(biomaRt)
library(survival)
library(RegParallel)
library(survminer)
library(matrixStats)
library(ggplot2)
library(ggrepel)
require(statar) # for xtile
library(GEOquery)
library(matrixStats)
library(data.table)
library(reshape2)

```

Set directories.
```{r}

dir.output <- "~/datasets/SLpediatric/sl_pairs_results/"
dir.figures <- "~/Google Drive/NIH_NCI/Projects/AML_pediatric/figures/"
dir.code <- "~/datasets/SLpediatric/R/"
dir.repo <- "~/src/SLpediatric/R/"
dir.rdata <- "~/datasets/SLpediatric/Rdata_object/version6_nbl1-3/"
dir.ppi <- "~/datasets/STRING/"
dir.depmap <- "~/datasets/DepMap/"
dir.aml <- "~/datasets/SLpediatric/Rdata_object/version7_aml/"
dir.target <- "~/datasets/SLpediatric/datasets/TARGET/"
dir.beat <- "~/datasets/AML/Tyner_etal_BEATstudy/"
dir.save <- "~/datasets/AML/result_tables/"

```

Functions for subsetting prob Rdata objects.
```{r}
subset.prob.tcga <- function(prob, ix){
  prob$samples <- prob$samples[ix]
  prob$types <- prob$types[ix]
  prob$mRNA <- prob$mRNA[,ix]
  prob$scna <- prob$scna[,ix]
  prob$mRNAq <- prob$mRNAq[,ix]
  prob$scnaq <- prob$scnaq[,ix]
  prob$mRNAq2 <- prob$mRNAq2[,ix]
  prob$scnaq2 <- prob$scnaq2[,ix]
  prob$sex <- prob$sex[ix]
  prob$age <- prob$age[ix]
  prob$race <- prob$race[ix]
  prob$survival <- prob$survival[ix,]
  prob$stage <- prob$stage[ix]
  prob$mRNA.norm <- prob$mRNA.norm[,ix]
  prob$scna.norm <- prob$scna.norm[,ix]
  prob$surv.dt <- prob$surv.dt[ix,]
  prob$mrna.tpm <- prob$mrna.tpm[,ix]
  return(prob)
}

subset.genes.prob.tcga <- function(prob, ix){
  prob$genes <- prob$genes[ix]
  prob$entrez <- prob$entrez[ix]
  prob$samples <- prob$samples
  prob$types <- prob$types
  prob$mRNA <- prob$mRNA[ix, ]
  prob$scna <- prob$scna[ix, ]
  prob$mRNAq <- prob$mRNAq[ix, ]
  prob$scnaq <- prob$scnaq[ix, ]
  prob$mRNAq2 <- prob$mRNAq2[ix, ]
  prob$scnaq2 <- prob$scnaq2[ix, ]
  prob$sex <- prob$sex
  prob$age <- prob$age
  prob$race <- prob$race
  prob$survival <- prob$survival
  prob$stage <- prob$stage
  prob$mRNA.norm <- prob$mRNA.norm[ix, ]
  prob$scna.norm <- prob$scna.norm[ix, ]
  prob$surv.dt <- prob$surv.dt
  prob$mrna.tpm <- prob$mrna.tpm[ix, ]
  return(prob)
}

subset.genes.prob.target <- function(prob, ix){
  prob$genes <- prob$genes[ix]
  prob$ensembl <- prob$ensembl[ix]
  prob$ensembl_version <- prob$ensembl_version[ix]
  prob$scna <- prob$scna[ix, ]
  prob$mRNA <- prob$mRNA[ix, ]
  prob$mrna.tpm <- prob$mrna.tpm[ix,]
  prob$samples <- prob$samples
  prob$types <- prob$types
  prob$mutations <- prob$mutations
  prob$sex <- prob$sex
  prob$age <- prob$age
  prob$race <- prob$race
  prob$ethnicity <- prob$ethnicity
  prob$survival <- prob$survival
  prob$eventfree <- prob$eventfree
  return(prob) 
}

subset.genes.prob.beat <- function(prob, ix){
  prob$genes <- prob$genes[ix]
  prob$samples <- prob$samples
  prob$types <- prob$types
  prob$mrna.tpm <- prob$mrna.tpm[ix, ]
  prob$mrna.rpkm <- prob$mrna.rpkm[ix, ]
  prob$sex <- prob$sex
  prob$age <- prob$age
  prob$ethnicity <- prob$ethnicity
  prob$tissue <- prob$tissue
  prob$blastsinBM <- prob$blastsinBM
  prob$survival <- prob$survival
  prob$surv.data <- prob$surv.dat
  return(prob) 
}

```

Convert RPKM/FPKM to TPM
```{r}
# Implementation was taken from https://haroldpimentel.wordpress.com/2014/05/08/what-the-fpkm-a-review-rna-seq-expression-units/
fpkmToTpm <- function(fpkm){
  return(exp(log(fpkm) - log(sum(fpkm)) + log(1e6)))
}
```

Cox function.
```{r}
cox.gene.expression <- function(gene, surv_mat=surv_mat){
  
  cox_uncntrl <- coxph(Surv(time, status) ~ gene, data=surv_mat)
  summary_uncntrl  <- summary(cox_uncntrl)
    
  cox_cntrl <- coxph(Surv(time, status) ~ gene + age + strata(sex, race), data=surv_mat)
  summary_cntrl  <- summary(cox_cntrl)
    
  res <- c(summary_uncntrl$coefficients["gene", 1:5], summary_cntrl$coefficients["gene", 1:5])
  
  return(res)
}

cox.gene.expression.target <- function(gene, surv_mat=surv_mat){
  
  cox_uncntrl <- coxph(Surv(time, status) ~ gene, data=surv_mat)
  summary_uncntrl  <- summary(cox_uncntrl)
    
  cox_cntrl <- coxph(Surv(time, status) ~ gene + age + strata(sex, race, ethnicity, tissue), data=surv_mat)
  summary_cntrl  <- summary(cox_cntrl)
    
  res <- c(summary_uncntrl$coefficients["gene", 1:5], summary_cntrl$coefficients["gene", 1:5])
  
  return(res)
}

cox.gene.expression.beat <- function(gene, surv_mat=surv_mat){
  
  cox_uncntrl <- coxph(Surv(time, status) ~ gene, data=surv_mat)
  summary_uncntrl  <- summary(cox_uncntrl)
    
  cox_cntrl <- coxph(Surv(time, status) ~ gene + age + blastsinBM + strata(sex, ethnicity), data=surv_mat)
  summary_cntrl  <- summary(cox_cntrl)
  
  cox_cntrl_noblasts <- coxph(Surv(time, status) ~ gene + age + strata(sex, ethnicity), data=surv_mat)
  summary_cntrl_noblasts <- summary(cox_cntrl_noblasts)
    
  res <- c(summary_uncntrl$coefficients["gene", 1:5],
           summary_cntrl_noblasts$coefficients["gene", 1:5],
           summary_cntrl$coefficients["gene", 1:5]
           )
  
  return(res)
}

```

Load data.
```{r}

# load depmap data
load(paste(dir.depmap, "depmap_19Q3.RData", sep=""))
# load TCGA data
load(paste(dir.code, "prob.TCGA.extended.qnorm.RData", sep=""));prob0=prob
rm(prob)
# load TARGET data
load(paste(dir.aml, "aml.fpkms.n145.hg38.prob.Rdata", sep="")) # prob_name=aml.fpkms.prob.hg38
# read in data from BEAT study
#rna.beat.cpm <- read.csv(paste(dir.beat, "Gene_counts_CPM.csv", sep=""), header=TRUE, sep=",", stringsAsFactors = FALSE)
rna.beat.rpkm <- read.csv(paste(dir.beat, "Gene_counts_RPKM.csv", sep=""), header=TRUE, sep=",", stringsAsFactors = FALSE)
rna.beat.clinical <- read.csv(paste(dir.beat, "Clinical_Summary.csv", sep=""), header=TRUE, sep=",", stringsAsFactors = FALSE)
# load BEAT AML - TPM normalized
load(file=paste(dir.beat, "prob_aml_beat.RData", sep=""))
# Load drug data
load(paste(dir.code, "prob.CTRP.RData", sep=""));prob.ctrp=prob # prob.ctrp
# dont use it's faulty
#load(paste(dir.code, "prob.CTRP.v2.extended.RData", sep=""));prob.ctrp=prob # prob.ctrp
load(paste(dir.code, "prob.gdsc.new.extended.RData", sep=""))# prob.gdsc

```

Create Rdata object for BEAT AML study.
```{r}
# Convert log(RPKM) to RPKM
rna.beat <- exp(as.matrix(rna.beat.rpkm[, 3:dim(rna.beat.rpkm)[2]]))
# Convert RPKM to TPM
rna.tpm <- apply(rna.beat, 2, function(x) fpkmToTpm(x))

# Issue
# Readin through the methods section of the BEAT study, RPKM values were calculated as proposed by
# Hansen, K. D., Irizarry, R. A. & Wu, Z. Removing technical variability in RNA-seq data using 
# conditional quantile normalization. Biostatistics 13, 204–216 (2012).
# Tyner et al., "Normalization was performed using the conditional quantile normalization 
# procedure (64), which produced GC-content-corrected log2 reads per kilobase per million 
# mapped reads (RPKM) values. This procedure produces both offsets to be used in conjunction 
# with edgeR as well as a matrix of log2-normalized RPKM values for clustering"

# rename column names in AML cohort
colnames(rna.tpm) <- gsub("\\.", "-", gsub("X", "AML-", colnames(rna.tpm)))
colnames(rna.beat.rpkm) <- gsub("\\.", "-", gsub("X", "AML-", colnames(rna.beat.rpkm)))
# rename row names in AML cohort
row.names(rna.tpm) <- rna.beat.rpkm$Symbol
row.names(rna.beat.rpkm) <- rna.beat.rpkm$Symbol

# change LabId name to match AML cohort
rna.beat.clinical$LabId_match_with_rpkm_matrix <- paste("AML-", rna.beat.clinical$LabId, sep="")

# only keep Labids that match with AML cohort
rna.beat.clinical.sel1 <- rna.beat.clinical[rna.beat.clinical$LabId_match_with_rpkm_matrix %in% colnames(rna.tpm), ]
rna.beat.clinical.sel1 <- rna.beat.clinical[rna.beat.clinical$LabId_match_with_rpkm_matrix %in% colnames(rna.beat.rpkm), ]

# select cases with the following specimenGroup
sel.specimenGroups <- c("Initial Acute Leukemia Diagnosis", 
                       "Post-Chemotherapy", "Post-Chemotherapy|Relapse",
                       "Post-Chemotherapy|Residual Disease", "Residual Disease",
                       "Residual Disease|Post-Chemotherapy")
# remove 'Post-Chemotherapy'
afterPCAsel <- c("Initial Acute Leukemia Diagnosis", "Post-Chemotherapy|Relapse",
                 "Post-Chemotherapy|Residual Disease", "Residual Disease",
                 "Residual Disease|Post-Chemotherapy")
# select samples 
rna.beat.clinical.sel2 <- rna.beat.clinical.sel1[rna.beat.clinical.sel1$specimenGroups %in% afterPCAsel, ]

# select samples from rna matrix
rna.tpm.sel <- rna.tpm[, rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix]
rna.beat.rpkm.sel <- rna.beat.rpkm[,rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix]

# sort clinical according to rna matrix rna.tpm
rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix <- factor(rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix, 
                                                              levels=colnames(rna.tpm.sel))
rna.beat.clinical.sel2 <- rna.beat.clinical.sel2[order(rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix), ]
# sort clinical to rna matrix rna rpkm
rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix <- factor(rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix, 
                                                              levels=colnames(rna.beat.rpkm.sel))
rna.beat.clinical.sel2 <- rna.beat.clinical.sel2[order(rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix), ]

# survival data
rna.beat.clinical.sel2$status <- 1
rna.beat.clinical.sel2[rna.beat.clinical.sel2$vitalStatus=="Dead",]$status <- 0
survival <- data.table(
  data.frame(
    time=rna.beat.clinical.sel2$overallSurvival,
    status=rna.beat.clinical.sel2$status))

# change codex for survival (death=1, alive=0), needed for cox regression
surv.dat <- data.frame(time=survival[,1], status=ifelse(survival[,2]==0,1,0))


# create Rdata object
aml_beat <- list(
  genes=as.character(row.names(rna.tpm.sel)),
  samples=as.character(colnames(rna.tpm.sel)),
  types=rep("BEAT-AML", times=length(colnames(rna.tpm.sel))),
  mrna.rpkm=as.matrix(rna.beat.rpkm.sel), 
  mrna.tpm=rna.tpm.sel,
  sex=as.character(toupper(rna.beat.clinical.sel2$consensus_sex)),
  age=as.numeric(rna.beat.clinical.sel2$ageAtDiagnosis),
  ethnicity=as.character(toupper(rna.beat.clinical.sel2$inferred_ethnicity)),
  tissue=as.character(rna.beat.clinical.sel2$specimenType),
  blastsinBM=as.numeric(rna.beat.clinical.sel2$X..Blasts.in.BM),
  survival=survival,
  surv.data=surv.dat
  )

# save Robject
save(file=paste(dir.beat, "prob_aml_beat.RData", sep=""), aml_beat)

```

Plot PCA (For BEAT AML)
```{r}

mat2pca <- function(pca.matrix){
  # pca.matrix <- prob.nbl$mRNA
  pca.matrix <- as.matrix(pca.matrix)
  pca.compl.cases <- pca.matrix[complete.cases(pca.matrix), ]
  pca.compl.cases <- cbind(pca.compl.cases, rowMedians=rowMedians(pca.compl.cases))
  quantiles.pca <- quantile(pca.compl.cases[, "rowMedians"])
  pca.compl.cases <- as.data.frame(pca.compl.cases)
  pca.rm.low.expr <- pca.compl.cases[pca.compl.cases$rowMedians > quantiles.pca[2], ]
  pca.rm.low.expr.sel <- pca.rm.low.expr[, 1:(dim(pca.rm.low.expr)[2]-1)]
  pca.res <- prcomp(pca.rm.low.expr.sel)
  pcas <- pca.res$rotation
  plot.pca <- data.frame(PC1=pcas[,c("PC1")], PC2=pcas[,c("PC2")])
  return(pcas)
}

#plot.pca <- function(plot.fig.pca, VAR="isRelapse"){
#  g <- ggplot(data=plot.fig.pca, aes(x=PC1, y=PC2, colour=specimenGroups))
#  g <- g + geom_point(size=2)
#  #g <- g + ggtitle(title)
#  g <- g + theme_bw()
#  plot(g)
#  return(g)
#} 

# preprocess matrix
# subselect rpkm matrix
rna.beat.rpkm.sel <- rna.beat.rpkm[, rna.beat.clinical.sel2$LabId_match_with_rpkm_matrix]
pcas <- mat2pca(rna.beat.rpkm.sel) # add variable to plot
pcas <- as.data.frame(pcas)
pcas$LabId_match_with_rpkm_matrix <- rownames(pcas)
m.pcas <- merge(pcas[, c("PC1", "PC2", "LabId_match_with_rpkm_matrix")], rna.beat.clinical.sel2, by="LabId_match_with_rpkm_matrix")
m.pcas$X..Blasts.in.BM <- as.numeric(m.pcas$X..Blasts.in.BM)
m.pcas$X..Blasts.in.PB <- as.numeric(m.pcas$X..Blasts.in.PB)
m.pcas$Platelet.Count <- as.numeric(m.pcas$Platelet.Count)
m.pcas$WBC.Count<- as.numeric(m.pcas$WBC.Count)
m.pcas$X..Immature.Granulocytes.in.PB <- as.numeric(m.pcas$X..Immature.Granulocytes.in.PB)

# plots
# specimenGroups, X..Blasts.in.BM --> these variables show distinct clustering or gradient
g <- ggplot(data=m.pcas, 
            aes(x=PC1, y=PC2, colour=ELN2017))
g <- g + geom_point(size=2, alpha=0.7)
#g <- g + scale_color_gradient(low="blue", high="red")
g <- g + theme_bw()
plot(g)

```

Extract AML dataset from GEO (GSE67040).
```{r}
# Requested survival data.
aml.rna.geoquery <- getGEO("GSE67040")

```

Select Ensemble mirror.
```{r}
# Select ensembl mirror
ensembl <- useEnsembl(
  biomart="ensembl", 
  dataset="hsapiens_gene_ensembl",
  mirror="useast")
```

Preprocess TCGA data.
```{r}
# select LAML only
ix <- which(prob0$types=="LAML")
prob_adult_aml <- subset.prob.tcga(prob0, ix)

# free up some memory
rm(prob0)

# Convert FPKM to TPM
mrna.tpm <- apply(prob_adult_aml$mRNA, 2, function(x) fpkmToTpm(x))
prob_adult_aml$mrna.tpm <- mrna.tpm

# select only protein coding genes
gene_biotype <- getBM(
  attributes = c('hgnc_symbol', 'gene_biotype'), 
  filters = 'hgnc_symbol', 
  values = prob_adult_aml$genes, 
  mart = ensembl)

# index of all protein coding genes
all_protein_coding <- gene_biotype[gene_biotype$gene_biotype=="protein_coding", ]$hgnc_symbol
ix <- which(prob_adult_aml$genes %in% all_protein_coding)

# subset matrix
aml_intermediate <- subset.genes.prob.tcga(prob_adult_aml, ix)

# only select genes with a median TPM >1
ix <- which(rowMedians(aml_intermediate$mrna.tpm) > 1)
aml_tcga <- subset.genes.prob.tcga(aml_intermediate, ix)

# assign column and rownames to mRNA
colnames(aml_tcga$mrna.tpm) <- aml_tcga$samples
row.names(aml_tcga$mrna.tpm) <- aml_tcga$genes

```

Preprocess TARGET data.
```{r}

# Convert FPKM to TPM
mrna.tpm <- apply(aml.fpkms.prob.hg38$mRNA, 2, function(x) fpkmToTpm(x))
aml.fpkms.prob.hg38$mrna.tpm <- mrna.tpm

# only select protein coding genes
aml.fpkms.prob.hg38$ensembl_version <- aml.fpkms.prob.hg38$ensembl
aml.fpkms.prob.hg38$ensembl <- sapply(
  aml.fpkms.prob.hg38$ensembl_version, 
  function(x) strsplit(as.character(x), split="\\.")[[1]][1])
row.names(aml.fpkms.prob.hg38$mRNA) <- aml.fpkms.prob.hg38$ensembl
row.names(aml.fpkms.prob.hg38$mrna.tpm) <- aml.fpkms.prob.hg38$ensembl

# select only protein coding genes
gene_biotype <- getBM(
  attributes = c('ensembl_gene_id', 'gene_biotype'), 
  filters = 'ensembl_gene_id', 
  values = aml.fpkms.prob.hg38$ensembl, 
  mart = ensembl)

# index of all protein coding genes
all_protein_coding <- gene_biotype[gene_biotype$gene_biotype=="protein_coding", ]$ensembl_gene_id
ix <- which(aml.fpkms.prob.hg38$ensembl %in% all_protein_coding)

# subset matrix
pediatric_aml_intermediate <- subset.genes.prob.target(aml.fpkms.prob.hg38, ix)

# only select genes with a median TPM >1
ix <- which(rowMedians(pediatric_aml_intermediate$mrna.tpm) > 1)
pediatric_aml <- subset.genes.prob.target(pediatric_aml_intermediate, ix)

# change codex for survival (death=1, alive=0), needed for cox regression
pediatric_aml$surv.data <-  data.frame(time=pediatric_aml$survival[,1], 
           status=ifelse(pediatric_aml$survival[,2]==0,1,0))

```

Preprocess BEAT AML data.
```{r}
# only select protein coding genes
gene_biotype <- getBM(
  attributes = c('hgnc_symbol', 'gene_biotype'), 
  filters = 'hgnc_symbol', 
  values = aml_beat$genes, 
  mart = ensembl)

# index of all protein coding genes
all_protein_coding <- gene_biotype[gene_biotype$gene_biotype=="protein_coding", ]$hgnc_symbol
ix <- which(aml_beat$genes %in% all_protein_coding)

# subset matrix
aml_beat_intermediate <- subset.genes.prob.beat(aml_beat, ix)

# only select genes with a TPM greater than 1
ix <- which(rowMedians(as.matrix(aml_beat_intermediate$mrna.tpm)) > 1)
aml_beat_filtered <- subset.genes.prob.beat(aml_beat_intermediate, ix)

# only select genes with an FPKM greater than 0 fplm=log2(1=0)
ix <- which(rowMedians(as.matrix(aml_beat_intermediate$mrna.rpkm)) > 0)
aml_beat_filtered <- subset.genes.prob.beat(aml_beat_intermediate, ix)
```

Prognostic genes from TCGA.
```{r}

# create dataframe
surv_mat <- data.frame(
  samples=aml_tcga$samples,
  sex=aml_tcga$sex,
  age=aml_tcga$age,
  race=aml_tcga$race,
  stage=aml_tcga$stage,
  time=aml_tcga$surv.dt$time,
  status=aml_tcga$surv.dt$status)

# merge and transform data
# take log2 of FPKM values
cox_data <- cbind(surv_mat, t(log2(aml_tcga$mRNA + 1)))
scaled_data <- cbind(surv_mat, scale(t(log2(aml_tcga$mRNA + 1))))

##############################################################
# Run Cox Surv(time, status) ~ [*] + age + strata(sex, race) #
##############################################################
cox_mrna <- t(apply(t(log2(aml_tcga$mrna.tpm + 1)), 2, function(gene) cox.gene.expression(gene, surv_mat)))
colnames(cox_mrna) <- c("uncntrl_coef", "uncntrl_HR", "uncntrl_se_coeff", "uncntrl_z", "uncntrl_P",
                        "cntrl_coef", "cntrl_HR", "cntrl_se_coeff", "cntrl_z", "cntrl_P")
cox_mrna <- as.data.frame(cox_mrna)
cox_mrna$uncntrl_padjust <- p.adjust(cox_mrna$uncntrl_P, method="BH")
cox_mrna$cntrl_padjust <- p.adjust(cox_mrna$cntrl_P, method="BH")

# save results
# select all with a cntrl pvalue <0.2
cox_mrna_0p2 <- cox_mrna[cox_mrna$cntrl_padjust < 0.2, ] 
# annotate top genes with biomart
gene_biotype <- getBM(
  attributes = c('hgnc_symbol', 'description'), 
  filters = 'hgnc_symbol', 
  values = rownames(cox_mrna_0p2), 
  mart = ensembl)
# add gene names
cox_mrna_0p2$hgnc_symbol <- rownames(cox_mrna_0p2) 
# merge
cox_mrna_merge <- merge(cox_mrna_0p2, gene_biotype, by="hgnc_symbol", all.x=TRUE )
# sort by pvalue
cox_mrna_merge <- cox_mrna_merge[order(cox_mrna_merge$cntrl_padjust), ]
# write table
write.csv(cox_mrna_merge, file=paste(dir.save, "cox_tcga_tpm_results_with_description.csv", sep="/"), row.names=TRUE)

############################################################################
# Run Cox with scaled data, standardized each gene expression distribution #
############################################################################
scaled_data <- scale(t(log2(aml_tcga$mrna.tpm + 1)))
cox_mrna_scaled <- t(apply(scaled_data, 2, function(gene) cox.gene.expression(gene, surv_mat)))
colnames(cox_mrna_scaled) <- c("uncntrl_coef", "uncntrl_HR", "uncntrl_se_coeff", "uncntrl_z", "uncntrl_P",
                        "cntrl_coef", "cntrl_HR", "cntrl_se_coeff", "cntrl_z", "cntrl_P")
cox_mrna_scaled <- as.data.frame(cox_mrna_scaled)
cox_mrna_scaled$cntrl_padjust <- p.adjust(cox_mrna_scaled$cntrl_P, method="BH")
cox_mrna_scaled$uncntrl_padjust <- p.adjust(cox_mrna_scaled$uncntrl_P, method="BH")

# save results
# select all with a cntrl pvalue <0.2
cox_mrna_scaled_0p2 <- cox_mrna_scaled[cox_mrna_scaled$cntrl_padjust < 0.2, ] 
# annotate top genes with biomart
gene_biotype <- getBM(
  attributes = c('hgnc_symbol', 'description'), 
  filters = 'hgnc_symbol', 
  values = rownames(cox_mrna_scaled_0p2), 
  mart = ensembl)
# add gene names
cox_mrna_scaled_0p2$hgnc_symbol <- rownames(cox_mrna_scaled_0p2) 
# merge
cox_mrna_scaled_merge <- merge(cox_mrna_scaled_0p2, gene_biotype, by="hgnc_symbol", all.x=TRUE )
# sort by pvalue
cox_mrna_scaled_merge <- cox_mrna_scaled_merge[order(cox_mrna_scaled_merge$cntrl_padjust), ]
# write table
write.csv(cox_mrna_scaled_merge, file=paste(dir.save, "cox_tcga_scaled_tpm_results_with_description.csv", sep="/"), row.names=TRUE)

################################ 
# Plot Hazard ratio vs P value #
################################
# for plotting
cox_mrna <- cox_mrna_scaled
cox_mrna$hgnc_symbol <- rownames(cox_mrna)

p <- ggplot(data=cox_mrna, aes(x=log(cntrl_HR), y=-log10(cntrl_P)))
p <- p + geom_point(fill="blue", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_point(data=cox_mrna[cox_mrna$cntrl_padjust < 0.05, ], 
                    aes(x=log(cntrl_HR), y=-log10(cntrl_P)),
                    fill="red", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_text_repel(data=cox_mrna[cox_mrna$cntrl_padjust < 0.05 & log2(cox_mrna$cntrl_HR) > 0 , ], 
                         aes(x=log(cntrl_HR), y=-log10(cntrl_P), label=hgnc_symbol), 
                         color="red")
p <- p + geom_text_repel(data=cox_mrna[cox_mrna$cntrl_padjust < 0.05 & log2(cox_mrna$cntrl_HR) < 0 , ], 
                         aes(x=log(cntrl_HR), y=-log10(cntrl_P), label=hgnc_symbol), 
                         color="blue")
#p <- p + coord_fixed(ratio = 1)
p <- p + theme_bw()
p <- p + theme(axis.text.x=element_text(colour="black", size=10), 
               axis.text.y=element_text(colour="black", size=10))
print(p)

fn <- sprintf(paste(dir.figures, "cox_tcga_scaled_tpm_cntrl_fdr0p5.pdf", sep=""))
pdf(fn, height=8, width=8)
print(p)
dev.off()
```

Prognositic genes for TARGET
```{r}

surv_mat <- data.frame(
  samples=pediatric_aml$samples,
  sex=pediatric_aml$sex,
  age=pediatric_aml$age,
  race=pediatric_aml$race,
  ethnicity=pediatric_aml$ethnicity,
  tissue=pediatric_aml$tissue,
  time=pediatric_aml$surv.data$time,
  status=pediatric_aml$surv.data$status)

# scale data
scaled_data <- scale(t(log2(pediatric_aml$mrna.tpm + 1 )))

# Run Cox 
cox_mrna <- t(apply(t(log2(pediatric_aml$mrna.tpm + 1 )), 2, function(gene) cox.gene.expression.target(gene, surv_mat)))
colnames(cox_mrna) <- c("uncntrl_coef", "uncntrl_HR", "uncntrl_se_coeff", "uncntrl_z", "uncntrl_P",
                        "cntrl_coef", "cntrl_HR", "cntrl_se_coeff", "cntrl_z", "cntrl_P")
cox_mrna <- as.data.frame(cox_mrna)
cox_mrna$uncntrl_padjust <- p.adjust(cox_mrna$uncntrl_P, method="BH")
cox_mrna$cntrl_padjust <- p.adjust(cox_mrna$cntrl_P, method="BH")
cox_mrna$ensembl_gene_id <- rownames(cox_mrna)

# annotate genes with biomart
gene_biotype <- getBM(
  attributes = c('ensembl_gene_id', 'hgnc_symbol', 'description'), 
  filters = 'ensembl_gene_id', 
  values = cox_mrna$ensembl_gene_id, 
  mart = ensembl)

# merge
cox_mrna_m <- merge(cox_mrna, gene_biotype, by="ensembl_gene_id", all.x=TRUE)

# sort by pvalue
cox_mrna_m <- cox_mrna_m[order(cox_mrna_m$cntrl_P), ]
# write table
write.csv(cox_mrna_m, file=paste(dir.save, "cox_TARGET_tpm_results_with_description.csv", sep="/"), row.names=TRUE)

# plotting
p <- ggplot(data=cox_mrna_m, aes(x=log2(cntrl_HR), y=-log10(cntrl_P)))
p <- p + geom_point(fill="blue", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_point(data=cox_mrna_me[cox_mrna_m$cntrl_padjust < 0.05, ], 
                    aes(x=log2(cntrl_HR), y=-log10(cntrl_P)),
                    fill="red", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_text_repel(data=cox_mrna_m[cox_mrna_m$cntrl_padjust < 0.05 & log2(cox_mrna_m$cntrl_HR) > 0 , ], 
                         aes(x=log2(cntrl_HR), y=-log10(cntrl_P), label=hgnc_symbol), 
                         color="red")
p <- p + geom_text_repel(data=cox_mrna_m[cox_mrna_m$cntrl_padjust < 0.05 & log2(cox_mrna_m$cntrl_HR) < 0 , ], 
                         aes(x=log2(cntrl_HR), y=-log10(cntrl_P), label=hgnc_symbol), 
                         color="blue")
p <- p + coord_fixed(ratio = 1)
p <- p + theme_bw()
p <- p + theme(axis.text.x=element_text(colour="black", size=10), 
               axis.text.y=element_text(colour="black", size=10))
print(p)

fn <- sprintf(paste(dir.figures, "cox_target_tpm_cntrl_fdr0p5.pdf", sep=""))
pdf(fn, height=8, width=8)
print(p)
dev.off()
```

Prognostic genes for BEAT AML
```{r}

# create dataframe
surv_mat <- data.frame(
  samples=aml_beat_filtered$samples,
  sex=aml_beat_filtered$sex,
  age=aml_beat_filtered$age,
  ethnicity=aml_beat_filtered$ethnicity,
  tissue=aml_beat_filtered$tissue,
  blastsinBM=aml_beat_filtered$blastsinBM,
  time=aml_beat_filtered$surv.data$time,
  status=aml_beat_filtered$surv.data$status)

# Run Cox Surv(time, status) ~ [*] + age + strata(sex, ethnicity)
cox_mrna <- t(apply(t(log2(aml_beat_filtered$mrna.tpm + 1)), 2, function(gene) cox.gene.expression.beat(gene, surv_mat)))
# scaled version
scaled_data <- scale(t(log2(aml_beat_filtered$mrna.tpm + 1)))
cox_mrna <- t(apply(scaled_data, 2, function(gene) cox.gene.expression.beat(gene, surv_mat)))

# Run Cox Surv(time, status) ~ [*] + age + strata(sex, ethnicity) RPKM
cox_mrna <- t(apply(t(aml_beat_filtered$mrna.rpkm), 2, function(gene) cox.gene.expression.beat(gene, surv_mat)))
# scaled version
scaled_data <- scale(t(aml_beat_filtered$mrna.rpkm))
cox_mrna <- t(apply(scaled_data, 2, function(gene) cox.gene.expression.beat(gene, surv_mat)))

# rename column
colnames(cox_mrna) <- c("uncntrl_coef", "uncntrl_HR", "uncntrl_se_coeff", "uncntrl_z", "uncntrl_P",
                        "cntrl_noblast_coef", "cntrl_noblast_HR", "cntrl_noblast_se_coeff", "cntrl_noblast_z", "cntrl_noblast_P",
                        "cntrl_coef", "cntrl_HR", "cntrl_se_coeff", "cntrl_z", "cntrl_P")

cox_mrna <- as.data.frame(cox_mrna)
cox_mrna$cntrl_padjust <- p.adjust(cox_mrna$cntrl_P, method="BH")
cox_mrna$uncntrl_padjust <- p.adjust(cox_mrna$uncntrl_P, method="BH")
cox_mrna$cntrl_noblast_padjust <- p.adjust(cox_mrna$cntrl_noblast_P, method="BH")
cox_mrna$hgnc_symbol <- row.names(cox_mrna)

# save results
# write table
write.csv(cox_mrna, file=paste(dir.save, "cox_beat_tpm_results.csv", sep="/"), row.names=TRUE)
write.csv(cox_mrna, file=paste(dir.save, "cox_beat_rpkm_results.csv", sep="/"), row.names=TRUE)


# Plot Hazard ratio vs P value
p <- ggplot(data=cox_mrna, aes(x=log(cntrl_HR), y=-log10(cntrl_P)))
p <- p + geom_point(fill="blue", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_point(data=cox_mrna[cox_mrna$cntrl_padjust < 0.05, ], 
                    aes(x=log(cntrl_HR), y=-log10(cntrl_P)),
                    fill="red", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_text_repel(data=cox_mrna[cox_mrna$cntrl_padjust < 0.05 & log(cox_mrna$cntrl_HR) > 0 , ], 
                         aes(x=log(cntrl_HR), y=-log10(cntrl_P), label=hgnc_symbol), 
                         color="red")
p <- p + geom_text_repel(data=cox_mrna[cox_mrna$cntrl_padjust < 0.05 & log(cox_mrna$cntrl_HR) < 0 , ], 
                         aes(x=log(cntrl_HR), y=-log10(cntrl_P), label=hgnc_symbol), 
                         color="blue")
#p <- p + coord_fixed(ratio = 1)
p <- p + theme_bw()
p <- p + theme(axis.text.x=element_text(colour="black", size=10), 
               axis.text.y=element_text(colour="black", size=10))
print(p)

fn <- sprintf(paste(dir.figures, "cox_beat_rpkm_cntrl_fdr0p05.pdf", sep=""))
pdf(fn, height=6, width=8)
print(p)
dev.off()

# Plot Hazard ratio vs P value noblast
p <- ggplot(data=cox_mrna, aes(x=log(cntrl_noblast_HR), y=-log10(cntrl_noblast_P)))
p <- p + geom_point(fill="blue", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_point(data=cox_mrna[cox_mrna$cntrl_noblast_padjust < 0.05, ], 
                    aes(x=log(cntrl_noblast_HR), y=-log10(cntrl_noblast_P)),
                    fill="red", colour="white", pch=21, alpha=0.5, size=2.5)
p <- p + geom_text_repel(data=cox_mrna[cox_mrna$cntrl_noblast_padjust < 0.05 & log(cox_mrna$cntrl_noblast_HR) > 0 , ], 
                         aes(x=log(cntrl_noblast_HR), y=-log10(cntrl_noblast_P), label=hgnc_symbol), 
                         color="red")
p <- p + geom_text_repel(data=cox_mrna[cox_mrna$cntrl_noblast_padjust < 0.05 & log(cox_mrna$cntrl_noblast_HR) < 0 , ], 
                         aes(x=log(cntrl_noblast_HR), y=-log10(cntrl_noblast_P), label=hgnc_symbol), 
                         color="blue")
#p <- p + coord_fixed(ratio = 1)
p <- p + theme_bw()
p <- p + theme(axis.text.x=element_text(colour="black", size=10), 
               axis.text.y=element_text(colour="black", size=10))
print(p)

fn <- sprintf(paste(dir.figures, "cox_beat_rpkm_cntrl_noblast_fdr0p05.pdf", sep=""))
pdf(fn, height=6, width=8)
print(p)
dev.off()

```

Plot survival (Kaplan Maier plots).
```{r}
# try example with one gene

km.gene.obj <- function(cox_data, scaled_data=scaled_data, gene=gene, gene_symbol=gene_symbol){
  cox.g1 <- cox_data[, c("samples", "sex", "age", "race", "tissue", "time", "status", gene)]

  #survplotdata <- scaled_data[,c('time', 'status', gene)]
  #colnames(survplotdata) <- c('time', 'status', gene_symbol)

  # set Z-scale cut-offs for high and low expression
  #highExpr <- 1.0
  #lowExpr <- -0.5
  #survplotdata[[gene_symbol]] <- ifelse(survplotdata[[gene_symbol]] >= highExpr, 'High',
  #                        ifelse(survplotdata[[gene_symbol]] <= lowExpr, 'Low', 'Mid'))
  # relevel the factors to have mid as the ref level
  #survplotdata[[gene_symbol]] <- factor(survplotdata[[gene_symbol]] ,
  #                         levels = c('Mid', 'Low', 'High'))
  
  # different approach
  cox.g1[[gene_symbol]] <- xtile(cox.g1[[gene]], 3)

  return(cox.g1)
}

survplotdata <- km.gene.obj(
  cox_data=cox_data,
  scaled_data=scaled_data,
  gene="ENSG00000067048", 
  gene_symbol = "DDX3Y")

p <- ggsurvplot(
  survfit(Surv(time, status) ~ DDX3Y, data = survplotdata),
  data = survplotdata,
  risk.table = TRUE,
  pval = FALSE,
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE)

fn <- sprintf(paste(dir.figures, "kaplan_maier_DDX3Y.pdf", sep=""))
pdf(fn, height=6, width=6)
print(p)
dev.off()

```

Overlap significant genes with the cancer dependency map.
```{r}

# Q: Take the top ranked genes and compare their dependency in AML-specific vs non AML-specific cell lines
# Take the ranking of the cell lines....
# Note: Essentiality imputation: was removed

# Rank normalizaion function
rank.array <- function(mat){
  mat.back = mat 
  mat = mat[!is.na(mat)]
  mat = rank(mat, ties.method = "average")/length(mat);
  mat.back[!is.na(mat.back)] = mat
  mat.back
}

# Run wilcoxon
test.sig <- function(cox_mrna, depmap_19Q3_rank_norm, random=FALSE, rand.num=42){
  # select significant genes with adverse risk
  if(random){
    sig.genes <-  rownames(depmap_19Q3_rank_norm)[sample(1:nrow(depmap_19Q3_rank_norm), rand.num)]
    
  }
  else{
    sig.genes <- cox_mrna[cox_mrna$cntrl_noblast_padjust < 0.05 & 
                                  cox_mrna$cntrl_noblast_HR >1,]$hgnc_symbol
  }
  
  # Select genes to compare:
  sel.genes <- which(rownames(depmap_19Q3_rank_norm) %in% sig.genes)
  depmap_sel <- depmap_19Q3_rank_norm[sel.genes, ]
  
  # test using wilcoxon:
  wilcoxon_dep <- apply(depmap_sel, 1, function(x) wilcox.test(
  x[aml_cell_lines], x[non_aml_cl] , alternative='less')$p.value)
  
  return(wilcoxon_dep)
}


# Select AML cell lines: 
aml_cell_lines <- rownames(depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype=="AML", ])
non_aml_cl <- rownames(depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype!="AML", ])
adult_aml_cl <- rownames(depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype=="AML" & 
                                            depmap_19Q3$meta$age > 20, ])
adult_aml_cl <- adult_aml_cl[grepl('^ACH', adult_aml_cl)]

# Rank normalized essentiality matrix
depmap_19Q3_rank_norm <- apply(depmap_19Q3$avana, 2, rank.array)

# Run test once for significant genes
wilcox.pvalues <- test.sig(cox_mrna, depmap_19Q3_rank_norm, random=FALSE)

# How many are significant ?
table(wilcox.pvalues < 0.1) [2]

# Calculate empirical p-value:
# Empirical pvalues are calculated as p=r/n, where
# n is the number of replicate samples and r is the 
# number of these replicates that produce a test statistic 
# greater than or equal to that calculated for the actual data

wilcoxon.empirical.p <- function(repeat.n=100, p.threshold){
  for(i in 1:repeat.n){
    wilcox_pvalues <- test.sig(cox_mrna_merge, depmap_19Q3_rank_norm, random=TRUE)
    results[i, ] <- as.numeric(table(wilcox_pvalues < p.threshold))
  }
  return(results)
}

# run function for 0.1 threshold
results <- data.frame(false.events=numeric(), true.events=numeric())
results <- wilcoxon.empirical.p(repeat.n=1000, p.threshold=0.1)
# count empirical p
table(results$true.events >= as.numeric(table(wilcoxon_dep <= 0.1)[2]))/1000
# FALSE  TRUE 
# 0.935 0.065

# run function for 0.05 threshold
results <- data.frame(false.events=numeric(), true.events=numeric())
results <- wilcoxon.empirical.p(repeat.n=1000, p.threshold=0.05)
# count empirical p
table(results$true.events >= as.numeric(table(wilcoxon_dep < 0.05)[2]))/1000
# FALSE  TRUE 
# 0.562 0.438

# plot data
# wilcox.pvalues

generate.plot.df <- function(
  genes=c("CCND3"),
  aml_cell_lines,
  non_aml_cl,
  adult_aml_cl){
  
  dependency_score <- reshape2::melt(t(depmap_19Q3$avana[genes, ]))
  colnames(dependency_score) <- c("cl_id", "gene_name", "dependency_score")
  norm_rank <- reshape2::melt(t(depmap_19Q3_rank_norm[genes, ]))
  colnames(norm_rank) <- c("cl_id", "gene_name", "norm_rank")
  df_g <- cbind(dependency_score, norm_rank=norm_rank$norm_rank)
  # add grouping
  df_g$grouping <- "other_cell_line"
  df_g[df_g$cl_id %in% aml_cell_lines, ]$grouping <- "pediatric_AML_specific"
  df_g[df_g$cl_id %in% non_aml_cl, ]$grouping <- "non-AML_specific"
  df_g[df_g$cl_id %in% adult_aml_cl, ]$grouping <- "adult_AML_specific"
  
return(df_g)
}

# Plot 
selec <- names(which(wilcox.pvalues < 0.1))
selec <- names(wilcox.pvalues)
df_g <- generate.plot.df(genes=selec,
                         aml_cell_lines=aml_cell_lines, 
                         non_aml_cl=non_aml_cl, 
                         adult_aml_cl=adult_aml_cl)

# Give cell lines meaningful names
cl_name <- depmap_19Q3$meta[c("stripped_cell_line_name")]
cl_name$cl_id <- row.names(cl_name) 
df_gm <- merge(df_g, cl_name, by="cl_id")

g <- ggplot(df_gm, aes(x=dependency_score, y=norm_rank, fill=grouping))
g <- g + geom_point(
  pch=21, colour="black", fill="grey", size=3, alpha=0.8)
g <- g + geom_point(
  data=df_gm[df_gm$grouping=="adult_AML_specific", ], 
  aes(x=dependency_score, y=norm_rank), 
  pch=21, colour="black", fill="red", size=3, alpha=0.8)
g <- g + geom_point(
  data=df_gm[df_gm$grouping=="pediatric_AML_specific", ], 
  aes(x=dependency_score, y=norm_rank), 
  pch=21, colour="black", fill="yellow", size=3, alpha=0.8)
g <- g + geom_vline(xintercept= (-1), colour="red")
g <- g + geom_text_repel(
  data=df_gm[df_gm$dependency_score < (-1) & df_gm$grouping!="non-AML_specific", ], 
  aes(x=dependency_score, y=norm_rank, label=stripped_cell_line_name), color="blue")
g <- g + facet_wrap(~ gene_name)
plot(g)

# save plot
fn <- sprintf(paste(dir.figures, "depmap_BEAT_tpm_cntrl_fdr0p05_wilcox_all_gene_name.pdf", sep=""))
pdf(fn, height=8, width=12)
print(g)
dev.off()

```

Mine AML (adult & pediatric cell lines).
```{r}
# Validate prognostic genes with stat. significance in AML cell lines
# For TCGA select all 
# Q: Is the essentiality reduced in those cell lines with initial high expresssion of
# gene X and subsequent knock-down of gene X

# select only genes with p.adjust (wald.test) < 0.05
sig_genes_bad <- cox_mrna[cox_mrna$cntrl_padjust < 0.05 & cox_mrna$cntrl_HR > 1, ]
sig_genes_good <- cox_mrna[cox_mrna$cntrl_padjust < 0.05 & cox_mrna$cntrl_HR < 1, ]

# Essentiality imputation
for (i in seq(nrow(depmap_19Q3$avana))){
  median.na <- median(depmap_19Q3$avana[i,], na.rm=TRUE)
  if (is.na(median.na)) { median.na <- 0 }
  depmap_19Q3$avana[i, is.na(depmap_19Q3$avana[i,])] <- median.na
}

# selct adult AML cell lines
# had to take out age ....& depmap_19Q3$meta$age > 18
aml_df <- depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype=="AML" & depmap_19Q3$meta$age >18, ]
aml_adult <- row.names(aml_df[complete.cases(aml_df$age), ])
isel <- which(colnames(depmap_19Q3$avana) %in% aml_adult)
ava_id <- colnames(depmap_19Q3$avana[, isel])
mrna_isel <- depmap_19Q3$expression[, ava_id]

# set alpha
alpha=1/3

mrna1 <- mrna_isel > apply(mrna_isel, 1, quantile, probs = alpha,  na.rm = TRUE)
mrna2 <- mrna_isel > apply(mrna_isel, 1, quantile, probs = 1-alpha,  na.rm = TRUE)
depmap_19Q3$exprq <- mrna1 + mrna2

# select genes of interest
igsel <- depmap_19Q3$exprq[row.names(sig_genes_bad), ]
igsel <- depmap_19Q3$exprq[row.names(sig_genes_good), ]
    
# divide in low and high, perform wilcoxon
stat <- NULL
i <- 1
for (gene in row.names(sig_genes_good)){
  
  print(gene)
  ihi <- names(which((igsel[gene, ] == 2)))
  ilo <- names(which((igsel[gene, ] == 0)))
      
  if (length(ihi) > 1 & length(ilo) > 1 ){
      stat[i] <- wilcox.test(depmap_19Q3$avana[gene, ihi],
                              depmap_19Q3$avana[gene, ilo], 
                             alternative="greater")$p.value
  }
  
  i <- i + 1
}

df <- data.frame(gene=row.names(sig_genes_good), stat_wilcox=stat)

      
```

Mine drug response data.
```{r}
# Which genes are drug targets?

#TCGA genes
cox_res <- read.csv(file=paste(dir.save, "cox_tcga_tpm_results_with_description.csv", sep="/"))
sig.genes <- cox_res[cox_res$cntrl_padjust < 0.05 & cox_res$cntrl_HR > 1, ]

drug.target.list <- sapply(as.list(depmap_19Q3$drug_target_map$target), 
                           function(x) unlist(strsplit(gsub(" ", "", x, fixed = TRUE), split=",")))
annotate.drug <- function(drug.target.list=drug.target.list, gene.list=gene){
  drug.target.match.line <- c()
  for (gene in gene.list){
      dtm <- paste0(
        unique(depmap_19Q3$drug_target_map[
          which(
            sapply(
              drug.target.list, function(x) gene %in% x)),]$name), collapse=",")
      drug.target.match.line <- c(drug.target.match.line, dtm)
  }
  return(drug.target.match.line)
}
# method "single"
sig.genes$drugs <- annotate.drug(drug.target.list=drug.target.list, gene.list=as.character(sig.genes$hgnc_symbol))

# Q: Compare sensitivity between AML and non AML cell lines
drugs.to.test <- unlist(strsplit(paste(sig.genes[which(sig.genes$drugs != ""), ]$drugs, collapse = ","), ","))

ix <- which(depmap_19Q3$drug_target_map$name %in% drugs.to.test)
drugs_list <- depmap_19Q3$drug_target_map[which(depmap_19Q3$drug_target_map$name %in% drugs.to.test), ]

# Run each drug with wilcox
test.each.drug <- function(depmap_19Q3_rank_norm, drug_list){

  # Select genes to compare:
  sel_drugs <- which(rownames(depmap_19Q3_rank_norm) %in% drug_list)
  depmap_sel <- depmap_19Q3_rank_norm[sel_drugs, ]
  
  # test using wilcoxon:
  wilcoxon_dep <- apply(depmap_sel, 1, function(x) wilcox.test(
  x[aml_cell_lines], x[non_aml_cl] , alternative='less')$p.value)
  
  return(wilcoxon_dep)
}

# Rank normalize drug PRISM
depmap_19Q3_rank_norm <- apply(depmap_19Q3$drug_prism, 2, rank.array)
# try other dataset
row.names(prob.ctrp$mat) <- prob.ctrp$drugs
colnames(prob.ctrp$mat) <- prob.ctrp$celllines
ctrp_rank_norm <- apply(prob.ctrp$mat, 2, rank.array)

# only matching drug is bosutinib

# Run cox, ISSIE PRISM does not have a single AML cell line !!!!!
aml_cell_lines <- depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype=="AML", ]$stripped_cell_line_name
non_aml_cl <- depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype!="AML", ]$stripped_cell_line_name
aml_cell_lines <- aml_cell_lines[which(colnames(depmap_19Q3_rank_norm) %in% aml_cell_lines)]
non_aml_cl <- non_aml_cl[which(colnames(depmap_19Q3_rank_norm) %in% non_aml_cl)]
test.each.drug(depmap_19Q3_rank_norm, drug_list=drugs_list$broad_id_trimmed)

# same for CTRP
aml_cell_lines <- depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype=="AML", ]$stripped_cell_line_name
non_aml_cl <- depmap_19Q3$meta[depmap_19Q3$meta$lineage_subtype!="AML", ]$stripped_cell_line_name

amL_cell_lines <- aml_cell_lines[aml_cell_lines %in% colnames(ctrp_rank_norm)]
non_aml_cl <- non_aml_cl[non_aml_cl %in% colnames(ctrp_rank_norm)]

# select drugs
ctrp.matching.drugs <- drugs.to.test[which(drugs.to.test %in% rownames(ctrp_rank_norm))]
comp <- ctrp_rank_norm[ctrp.matching.drugs, ]

wilcox.test(comp[aml_cell_lines], comp[non_aml_cl] , alternative='less')

```

Independent validation.
```{r}
#TCGA genes
cox_res <- read.csv(file=paste(dir.save, "cox_tcga_tpm_results_with_description.csv", sep="/"))
sig.genes <- cox_res[cox_res$cntrl_padjust < 0.05 , ]
sig.genes$cntrl_log_HR <- log(sig.genes$cntrl_HR)

# IS IT PREDICTIVE WITHIN THE SAME COHORT (TCGA)?  CROSS VALIDATION
ix <- which(row.names(aml_tcga$mrna.tpm) %in% as.character(sig.genes$hgnc_symbol))
tcga_genes <- aml_tcga$mrna.tpm[ix, ]
tcga_genes <- log2(tcga_genes + 1)

sig.genes <- sig.genes[sig.genes$hgnc_symbol %in% row.names(tcga_genes), ]
sig.genes$hgnc_symbol <- factor(sig.genes$hgnc_symbol, levels=row.names(tcga_genes))
sig.genes <- na.omit(sig.genes[order(sig.genes$hgnc_symbol), ])
sig.genes$cntrl_log_HR <- log(sig.genes$cntrl_HR)

# Risk_score for each patient
weighted.hr <- apply(tcga_genes, 2, function(x) x*sig.genes$cntrl_log_HR)
risk_score_per_patient <- apply(weighted.hr, 2, sum)
ihi <- which(risk_score_per_patient > quantile(risk_score_per_patient, 0.5, na.rm=TRUE))
ilo <- which(risk_score_per_patient <= quantile(risk_score_per_patient, 0.5, na.rm=TRUE))

df <- data.frame(patient=names(risk_score_per_patient), 
                 risk_score=risk_score_per_patient,
                 grouping=xtile(risk_score_per_patient, 3))

#Compare survival between ihi and ilo
surv_mat <- data.frame(
  samples=aml_tcga$samples,
  sex=aml_tcga$sex,
  age=aml_tcga$age,
  race=aml_tcga$race,
  time=aml_tcga$surv.dt$time,
  status=aml_tcga$surv.dt$status)

surv_mat$strata <- "mid"
surv_mat[surv_mat$samples %in% as.character(df[df$grouping==3,]$patient), ]$strata <- "ihi"
surv_mat[surv_mat$samples %in% as.character(df[df$grouping==1,]$patient), ]$strata <- "ilo"

p <- ggsurvplot(
  survfit(Surv(time, status) ~ strata, data = surv_mat),
  data = surv_mat,
  risk.table = TRUE,
  pval = TRUE,
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE)

# BEAT COHORT
# for each patient calculate a risk score (in BEAT study)
ix <- which(row.names(aml_beat$mrna.tpm) %in% as.character(sig.genes$hgnc_symbol))
beat_genes <- aml_beat$mrna.tpm[ix, ]
beat_genes <- log2(beat_genes + 1)
sig.genes <- sig.genes[sig.genes$hgnc_symbol %in% row.names(beat_genes), ]
sig.genes$hgnc_symbol <- factor(sig.genes$hgnc_symbol, levels=row.names(beat_genes))
sig.genes <- na.omit(sig.genes[order(sig.genes$hgnc_symbol), ])
sig.genes$cntrl_log_HR <- log(sig.genes$cntrl_HR)
# Risk_score for each patient
weighted.hr <- apply(beat_genes, 2, function(x) x*sig.genes$cntrl_log_HR)
risk_score_per_patient <- apply(weighted.hr, 2, sum)
ihi <- which(risk_score_per_patient > quantile(risk_score_per_patient, 0.5, na.rm=TRUE))
ilo <- which(risk_score_per_patient <= quantile(risk_score_per_patient, 0.5, na.rm=TRUE))

df <- data.frame(patient=names(risk_score_per_patient), 
                 risk_score=risk_score_per_patient,
                 grouping=xtile(risk_score_per_patient, 3))

#Compare survival between ihi and ilo
surv_mat <- data.frame(
  samples=aml_beat$samples,
  sex=aml_beat$sex,
  age=aml_beat_filtered$age,
  ethnicity=aml_beat$ethnicity,
  tissue=aml_beat$tissue,
  blastsinBM=aml_beat$blastsinBM,
  time=aml_beat$surv.data$time,
  status=aml_beat$surv.data$status)

surv_mat$strata <- "mid"
surv_mat[surv_mat$samples %in% as.character(df[df$grouping==3,]$patient), ]$strata <- "ihi"
surv_mat[surv_mat$samples %in% as.character(df[df$grouping==1,]$patient), ]$strata <- "ilo"

p <- ggsurvplot(
  survfit(Surv(time, status) ~ strata, data = surv_mat),
  data = surv_mat,
  risk.table = TRUE,
  pval = TRUE,
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE)

fn <- sprintf(paste(dir.figures, "kaplan_maier_risk_score.pdf", sep=""))
pdf(fn, height=6, width=6)
print(p)
dev.off()

# REPEAT THE SAME IN TARGET COHORT
# add surv.data
aml.fpkms.prob.hg38$surv.data <- data.frame(
  time=aml.fpkms.prob.hg38$survival[,1], 
  status=ifelse(aml.fpkms.prob.hg38$survival[,2]==0,1,0))
# only select protein coding genes
row.names(aml.fpkms.prob.hg38$mrna.tpm) <- aml.fpkms.prob.hg38$genes
target_aml <- log2(aml.fpkms.prob.hg38$mrna.tpm + 1) 
ix <- which(row.names(target_aml) %in% as.character(sig.genes$hgnc_symbol))
target_genes <- target_aml[ix, ]
# romove last PRICKLE4
target_genes <- target_genes[1:(dim(target_genes)[1]-1), ]

sig.genes <- sig.genes[sig.genes$hgnc_symbol %in% row.names(target_genes), ]
sig.genes$hgnc_symbol <- factor(sig.genes$hgnc_symbol, levels=unique(row.names(target_genes)))
sig.genes <- na.omit(sig.genes[order(sig.genes$hgnc_symbol), ])
sig.genes$cntrl_log_HR <- log(sig.genes$cntrl_HR)
# Risk_score for each patient
weighted.hr <- apply(target_genes, 2, function(x) x*sig.genes$cntrl_log_HR)
risk_score_per_patient <- apply(weighted.hr, 2, sum)
ihi <- which(risk_score_per_patient > quantile(risk_score_per_patient, 0.5, na.rm=TRUE))
ilo <- which(risk_score_per_patient <= quantile(risk_score_per_patient, 0.5, na.rm=TRUE))

df <- data.frame(patient=names(risk_score_per_patient), 
                 risk_score=risk_score_per_patient,
                 grouping=xtile(risk_score_per_patient, 3))

#Compare survival between ihi and ilo
surv_mat <- data.frame(
  samples=aml.fpkms.prob.hg38$samples,
  sex=aml.fpkms.prob.hg38$sex,
  age=aml.fpkms.prob.hg38$age,
  ethnicity=aml.fpkms.prob.hg38$ethnicity,
  time=aml.fpkms.prob.hg38$surv.data$time,
  status=aml.fpkms.prob.hg38$surv.data$status)

surv_mat$strata <- "mid"
surv_mat[surv_mat$samples %in% as.character(df[df$grouping==3,]$patient), ]$strata <- "ihi"
surv_mat[surv_mat$samples %in% as.character(df[df$grouping==1,]$patient), ]$strata <- "ilo"

p <- ggsurvplot(
  survfit(Surv(time, status) ~ strata, data = surv_mat),
  data = surv_mat,
  risk.table = TRUE,
  pval = TRUE,
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE)

fn <- sprintf(paste(dir.figures, "kaplan_maier_risk_score.pdf", sep=""))
pdf(fn, height=6, width=6)
print(p)
dev.off()

```

Go enrichment of pathways ?
```{r}
```

Example.
```{r}
# Tutorial from https://www.biostars.org/p/344233/

```


